---
title: "dplyr"
author: "Cameron Jones"
date: 'Stat 240: Summer 2023'
output: html_document
---

_Disclaimer: This material was presented briefly at in-person discussions as a brief recap of the week's concepts during the Spring '23 iteration of Stat 240. It is not a comprehensive list on everything you need to know for this topic._

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
```

# Concept Review: dplyr

The built-in R dataframe "chickwts" contains weight measurements on 71 baby chickens after being fed one of six diets for a month and a half. 

```{r}
slice_sample(chickwts, n = 8)
```

# Concept: The Pipe (%>%) Operator

(helpful tip: keyboard shortcut for the pipe)
- Windows: ctrl+shift+M
- Mac: command+shift+M

```{r}
# What it does: Just passes the thing on the "left" into the thing on the "right" as the first argument
x = c(2, 3, 5, NA)

sd(x, na.rm = T)
x %>% sd(na.rm = T)
```

```{r}
# Motivation: Making your dplyr work flow easier to read and produce!
chickwts %>% 
  group_by(feed) %>% 
  summarize(avgWeight = mean(weight)) %>% 
  arrange(avgWeight)

# Everything you can do with the pipe, you can do without it... but it's ugly and difficult to write
arrange(summarize(group_by(chickwts, feed), avgWeight = mean(weight)), avgWeight)
```

```{r}
# All dplyr commands take a dataframe in, and spit a dataframe out; so they can be connected with the pipe!

# arrange spits out a (sorted) dataframe; that dataframe gets passed in as the first argument to filter

arrange(chickwts, weight) #%>% filter(feed == "horsebean")
arrange(chickwts, weight) %>% filter(feed == "horsebean")
```

# Concept: Filtering Rows and Columns with filter() and select()

## filter()

filter() reduces the dataframe to the set of ROWS which satisfy a given condition.

```{r}
chickwts %>% 
  filter(feed == "horsebean") # Double equals '==' tests for equality.

chickwts %>% 
  filter(weight > 320)

# Other (in)equality testers: <, <=, >=, >, %in% (coming up!)
```

## select()

select() reduces a dataframe to a set of specified COLUMNS.

```{r}
chickwts %>% 
  select(weight) %>% 
  head()
```


# Concept: Creating Summary Tables with group_by() and summarize()

group_by() and summarize() work together to reduce a dataframe of many rows down to a small dataframe which contains summary measures.

```{r}

# Summarize alone, with no groups, just returns a specified summary of the entire dataframe

chickwts %>% 
  summarize(avgWeight = mean(weight),
            stdDevWeight = sd(weight))

# When a group_by is specified before summarize, those summaries are provided within each group

chickwts %>% 
  group_by(feed) %>% 
  summarize(avgWeight = mean(weight),
            stdDevWeight = sd(weight))  

# Note the syntax within summarize: newColumnName = formula(existingColumn)

```

