---
title: "STAT 240: stringr"
author: "Cameron Jones"
date: "Spring 2024"
output: rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=T,message=F,warning=F)
library(tidyverse)
```

# Overview

## Learning Outcomes

* These lectures will teach you how to:
    - Interact with and manipulate string data using `stringr`
    - Read and write regular expressions at a basic level

## Preliminaries

- Files to download to `COURSE/lecture/unit06-stringr`
    - `week6-stringr.Rmd`
    - `week6-stringr.html`
    
- Files to download to `COURSE/data`:
    - `exoplanet-confirmed_2022-09-12.csv` (You likely already have this from Discussion 2)
    
# stringr: Simple Functions for Common String Operations

- We have seen **string** data often in this class, but we have always worked "around" it, never manipulating them or interacting with them directly.

- The base R functions are somewhat hard to remember, hard to use, and don't apply well to dataframes.

- `tidyverse` contains the package `stringr`, which contains a suite of helpful functions for common string operations, optimized for dataframes - we will explore these functions in this lecture series!

- See the cheatsheet for these [here](https://evoldyn.gitlab.io/evomics-2018/ref-sheets/R_strings.pdf).

- We will first see the `str_` functions' basic use, and then show their wide utility when combined with a sub-language called **regular expressions**.


## Data

- We create a simple data frame consisting of the built in names in R.
    - `colors()` creates a vector of color names
    - `seq_along()` creates a sequence from 1 to the length of the argument, which should be a vector.

```{r}
df_colors = tibble(
  index = seq_along(colors()),
  colors = colors())

df_colors
```

## The "Observer" Functions: str_detect, str_count, and str_length

> The "observer" functions, `str_detect`, `str_count`, and `str_length`, don't give you back any part of the string; they just observe the string and tell you something useful about it.

*Note: When I describe the types of string function, such as "observer", that's just how I think of it. It's not a technical term and you don't have to remember it, but if it helps, it helps!*

### str_detect

> `str_detect(string, pattern)` returns `TRUE` or `FALSE` indicating if `string` contains `pattern`.

* For all of these string function definitions, `string` can be a single string, or a vector of strings,

```{r}
# As a single string: does "watermelon" contain "melon"?
str_detect("watermelon", "melon")
```


```{r}
# Element-wise on a vector of strings; for each of the first six colors, do they contain "white"?
firstSixColors = head(colors())
firstSixColors

str_detect(firstSixColors, "white")
```

- Our string data is often embedded in a column in a dataframe, so we will use these commands within `dplyr` commands like `mutate()` and `filter()`.

- `str_detect()` is particularly helpful within `filter()` since it is returning a vector of `TRUE` and `FALSE`.

```{r}
df_colors %>% 
  filter(str_detect(colors, "white"))
```

### str_count

> `str_count(string, pattern)` returns an integer count of how many times `string` contains `pattern`.

* How many times does each color contain the letter "a"?

```{r}
df_colors %>% 
  mutate(howManyAs = str_count(colors, "a"))
```

* Which color has the most "e"'s?

```{r}
df_colors %>% 
  mutate(howManyEs = str_count(colors, "e")) %>% 
  slice_max(howManyEs, n = 1)
```

### str_length

  > `str_length(string)` returns the number of characters in `string`; its length.

* What are the 5+ longest color names?

```{r}
df_colors %>% 
  mutate(length = str_length(colors)) %>% 
  slice_max(length, n = 5)
```

* Of the 8-letter color names, what are the alphabetically first and last, and how many are there?

```{r}
df_colors %>% 
  filter(str_length(colors) == 8) %>% 
  summarize(first = min(colors), last = max(colors), n = n())
```

---

##### EXERCISE: Practice the "Observer" Functions

* This exercise will reinforce the observer string functions by having students practice on questions of intermediate difficulty. 

* Using `df_colors`, **write code** to answer the following questions:

* Of the color names which contain at least two instances of "s",
    - Hint: This is like the `str_length` last example, except for the `str_*` command itself.

```{r}
# Answer here
df_colors %>% 
  filter(str_count(colors, "s") >= 2) %>% 
  summarize(first = min(colors), last = max(colors))
```

* How many colors contain "ee"?
    - Hint: We have seen counting logical conditions before with `sum(is.na())`... 

```{r}
# Answer here
df_colors %>% 
  summarize(containsEE = sum(str_detect(colors, "ee")))
```

* What is the shortest color name (or names, if there are ties) that contains a "y"?

```{r}
# Answer here
df_colors %>% 
  filter(str_detect(colors, "y")) %>% 
  slice_min(str_length(colors), n = 1)
```

---

## The "Pull" Functions: str_extract and str_sub

> The "pull" functions, `str_extract` and `str_sub`, return a specific part out from within a string.

### str_extract

> `str_extract(string, pattern)` returns whatever part of `string` matched `pattern`, if any, or NA otherwise.

* This command seems a little silly when we just match a literal pattern:

```{r}
df_colors %>% 
  mutate(notInteresting = str_extract(colors, "white")) %>% 
  head()
```

* This is true of all `str_` commands, but particularly this one - it becomes much more helpful when we introduce regular expressions.

* Here's a sneak preview: the regular expression `"[aeiou]{2}"` stands for a vowel, and then another vowel, which is allowed but not required to be different.

*Note: in these notes, "vowel" will refer to just the letters a, e, i, o, and u; never y.*

```{r}
df_colors %>% 
  mutate(twoVowels = str_extract(colors, "[aeiou]{2}"))
```

### str_sub

> `str_sub(string, start, end)` returns a subsection of `string`, beginning with and including the letter in position `start`, and ending with and including the letter in position `end`.

* Basic use example: extract the subsection of the string including and between the third and fifth character.

```{r}
# Notice the behavior of an "out-of-bounds" index on pear; it doesn't freak out, it just stops when the string is over
str_sub(c("apple", "orange", "pear"), 3, 5)
```

* Extension: The use of negative numbers counts from the back of the string, no matter how long it is!

```{r}
# Start with the second character, end at the second-to-last character
str_sub(c("apple", "orange", "pear"), 2, -2)
```

* More complex use on a dataframe, or more specifically a vector within it

```{r}
# Notice that putting the same number in str_sub for start and end just extracts that one letter
df_colors %>% 
  mutate(firstLetter = str_sub(colors, 1, 1),
         lastLetter = str_sub(colors, -1, -1))
```

* Application: What's the most common second letter for R colors?

```{r}
df_colors %>% 
  mutate(secondLetter = str_sub(colors, 2, 2)) %>% 
  count(secondLetter)
```

* "r" is the most common second letter because there are 100 versions of "gray", and each of them exists with both spellings "gray" and "grey".

## The "Edit" Functions: str_c, str_replace, and str_to_*

> The "edit" functions, `str_c`, `str_replace`, and `str_to_*` all return a new modified string based on the old ones.

### str_c

> `str_c(arg1, arg2, ...)` takes many string arguments and combines them: either elementwise into a vector of strings, or into a single "collapsed" string.

* Recall: vector arithmetic is R's internal system for handling combinations of vectors of unequal sizes.
  - In shorter terms, it is the reason `c(1, 2, 3) + 5` ends up as `1+5, 2+5, 3+5`.


```{r}
# The default behavior: Combine two vectors of strings element-wise

fruits = c("apple", "orange", "pear")
veggies = c("broccoli", "green beans", "peas")

str_c(fruits, veggies)
```

* You can add the `sep` argument to control the separator (by default nothing, technically `""`) between the combined elements of each vector.

```{r}
str_c(fruits, veggies, sep = " and ")
```

* Additionally, if you add the `collapse` argument, the input vector of strings will be "collapsed" down to a single one. The value of the `collapse` argument will be inserted between each original element.
    - This is fundamentally VERY different than the above use.

```{r}
# Notice: this is just one string, not a vector
str_c(fruits, collapse = ", ")
```

* Here's an example on a dataframe:

```{r}
# rev() stands for reverse(), and just returns the "backwards" version of a vector.
df_colors_augmented = df_colors %>% 
  mutate(reverse_colors = rev(colors))

df_colors_augmented
```

```{r}
df_colors_augmented %>% 
  mutate(pair = str_c(colors, reverse_colors, sep = " or ")) %>% 
  relocate(pair)
```


### str_replace

> `str_replace(string, pattern, replacement)` replaces the first instance of `pattern` in `string` with `replacement`. `str_replace_all(string, pattern, replacement)` replaces all instances of `pattern` in `string` with `replacement`.

* Once again, this command will become more powerful when we know regular expressions.

```{r}
# Replace the first instance of "e" in "watermelon" with "X"
str_replace("watermelon", "e", "X")
```

```{r}
# Replace all instances of "e" in "watermelon" with "E"
str_replace_all("watermelon", "e", "X")
```

* Now, extending to a dataframe:

```{r}
df_colors %>% 
  mutate(renamed = str_replace(colors, "antique", "vintage"))
```

* With regular expressions we can do operations such as remove any two consecutive vowels:

```{r}
df_colors %>% 
  mutate(noVowelPairs = str_replace(colors, "[aeiou]{2}", ""))
```

#### Shortcut: str_remove

* Getting rid of a certain `pattern` within a `string` is a common operation, so instead of typing `str_replace(string, pattern, replacement = "")` or `str_replace_all`, `stringr` gives you the shortcuts `str_remove` and `str_remove_all`.

```{r}
df_colors %>% 
  mutate(minusOneE = str_remove(colors, "e"),
         minusAllEs = str_remove_all(colors, "e"))
```

### str_to*

> The four commands `str_to_upper`, `str_to_lower`, `str_to_title`, and `str_to_sentence` change the **case** of the characters in the input string(s).

```{r}
myOpinion = "R is REALLY cool"

# casts all characters to lowercase
str_to_lower(myOpinion)

# casts all characters to uppercase
str_to_upper(myOpinion)

# casts the first letter of each word to uppercase, the rest to lowercase
str_to_title(myOpinion)

# casts the first letter of the string to uppercase, the rest to lowercase
str_to_sentence(myOpinion)
```

---

##### EXERCISE: Practice the "Pull" and "Edit" Functions

* This exercise will reinforce the "pull" and "edit" string functions by having students practice on questions of intermediate difficulty.

* Using `df_colors`, **write code** to answer the following questions:

* Display all the colors whose first letter is the same as their last letter.

```{r}
# Answer here
df_colors %>% 
  mutate(firstLetter = str_sub(colors, 1, 1),
         lastLetter = str_sub(colors, -1, -1)) %>% 
  filter(firstLetter == lastLetter)
```

* Create a new column, which contains the color names without any digits.

    - Hint 1: You don't have to remove each digit separately! Sneak preview... use the regular expression `"[:digit:]"` to stand for any digit!
    - Hint 2: Some numbers have multiple digits... we want to remove `all` of them.

```{r}
# Answer here
df_colors %>% 
  mutate(noNumbers = str_remove_all(colors, "[:digit:]"))
```

* Create a new column which contains the same exact color names, but those that contain the string "blue" are entirely in uppercase.
    - Hint: Use `case_when`!

```{r}
# Answer here
df_colors %>% 
  mutate(newColor = case_when(
    str_detect(colors, "blue") ~ str_to_upper(colors),
    TRUE ~ colors
  ))
```


---

# Regular Expressions

* Throughout these notes, we have hinted at and briefly seen regular expressions a few times; we now dive into them in more detail.

> **Regular expressions** are a language of simple characters used to capture complex string concepts.

* By "complex string concepts", we mean those that are easy to describe but hard or even impossible to check for manually with just literal characters.

* For example, regular expressions make checking for any two consecutive vowels very easy; we saw this earlier with `"[aeiou]{2}"`. 
    - However, with literal characters, we would have to check for "aa", "ae", "ai", "ao", "au", "ea"... et cetera.
    - You can see how this would quickly become unwieldy with even a slightly more complex pattern.

* Page 2 of [the stringr cheatsheet](https://evoldyn.gitlab.io/evomics-2018/ref-sheets/R_strings.pdf) contains a helpful reference.

## Data

* We're going to use a smaller, more manageable version of `df_colors` for the regular expression exercises.

```{r}
colors_small = df_colors %>% 
  slice(c(2, 8, 50, 130, 180, 210, 317, 452, 602, 652))

colors_small
```

## str_view

> `str_view(string, pattern)` is a helpful wrapper function as you learn regular expressions, that highlights exactly which part of `string` matches `pattern`. 

> `str_view` does not interact with dataframes, so you have to extract the vector manually.

* While we are learning how to use string operations and regular expressions within *dataframes*, this can be a complex context to learn them in.

* `str_view` helps you learn regular expressions, but requires that you are working with just a vector of strings, which we can extract with `pull` or the `$` operator.

* **We will just use `str_view` as an instructional tool during our introduction to regular expressions, it is not a practical command for data analysis.**

```{r}
# equivalent to: colors_small$colors
color_vector = colors_small %>% pull(colors)

color_vector
```

* By default, `str_view` shows you the elements in your vector which HAVE a match, and highlights the matching subsection:

```{r}
str_view(color_vector, "[:digit:]{2}")
```

* We will often set the argument `match = NA` which specifies to display ALL elements.

```{r}
str_view(color_vector, "[:digit:]{2}", match = NA)
```

## Sets and Alternatives

- `pattern1|pattern2|pattern3...` --- *or*, match pattern1 or pattern2 or pattern3, and so on
- `[abc]` --- *one of*, match one of "a" or "b" or "c"
- `[^abc]`--- *anything but*, match a character which is NOT one of "a" or "b" or "c"

- You can also use more complex ranges with `-`, such as... 
    - `[a-z]` for all lowercase letters
    - `[B-F]` for any uppercase letters between and including B and F
    - `[0-5]` for digits between and including 0 and 5
    - `[a-z0-5]` for any lowercase letter or any digit between and including 0 and 5.


* Which colors contain the pattern "gray" OR "grey"?

```{r}
str_view(color_vector, "gray|grey", match = NA)
```

* Here's another pattern which achieves the same goal:

```{r}
# stands literally for: a 'g', then an 'r', then an 'a' or an 'e', then a 'y'.
str_view(color_vector, "gr[ae]y", match = NA)
```

* Applied to our dataframe through `filter` and `str_detect`:

```{r}
colors_small %>% 
  filter(str_detect(colors, "gray|grey"))
```

* Which colors contain a vowel that is immediately followed by EITHER another vowel OR a digit?

```{r}
str_view(color_vector, "[aeiou][aeiou0-9]", match = NA)
```

### Sidenote: str_extract on multiple matches

* Notice that "chartreuse3" has multiple matches for the above pattern; "eu" AND "s3".

* `str_extract`, by default, will return the FIRST match.

```{r}
colors_small %>% 
  mutate(matching = str_extract(colors, "[aeiou][aeiou0-9]"))
```

* If we want to retain ALL matches, we can use `str_extract_all`, but be warned this data is un-tidy and may return a list-column.

* We don't teach you how to deal with list-columns formally in this course.

```{r}
colors_small %>% 
  mutate(matching = str_extract_all(colors, "[aeiou][aeiou0-9]"))
```

---

Back to examples!

* Which colors contain an "r" that is immediately followed by a consonant (a letter which is not a vowel)?
    - Technically, `[^aeiou]` includes numbers, symbols, spaces, and literally anything that is not a lowercase vowel - it happens to not matter here, but is a noteworthy cautionary tale!

```{r}
str_view(color_vector, "r[^aeiou]", match = NA)
```

```{r}
colors_small %>% 
  filter(str_detect(colors, "r[^aeiou]"))
```

## Anchors

- `^` --- the start of the string
    - This is the same symbol as the "anything but" set above; it acts as "anything but" WITHIN brackets, but outside of brackets it signifies the start of the string
- `$` --- the end of the string

## Quantifiers

- When these symbols appear AFTER a token, they modify how many times that token should match.

- `?` --- 0 or 1
- `*` --- 0 or more
- `+` --- 1 or more
- `{n}` --- exactly `n` times
- `{n,}` --- `n` or more
- `{m,n}` --- between (and including) `m` to `n` times

* Which colors START with two consonants?
    - Rather than list all consonants, let's just use "anything but" the vowels.

```{r}
# Trying taking the first carat out, and then both, and see how drastically the pattern changes!

str_view(color_vector, "^[^aeiou]{2}", match = NA)
```

```{r}
colors_small %>% 
  filter(str_detect(colors, "^[^aeiou]{2}"))
```

* Which colors contain the pattern: "An 'e', optionally followed by a single letter, followed by a single digit?

```{r}
str_view(color_vector, "e[a-z]?[0-9]", match = NA)
```

* Which colors END with the pattern described above?

```{r}
# Only addition in the pattern is the $ at the end, which stands for "end of string"
str_view(color_vector, "e[a-z]?[0-9]$", match = NA)
```


## Special Sets

- `.` --- matches literally any single character except for a newline, very commonly used   

- `[:space:]` --- space, tab, newline
- `[:blank:]` --- space and tab
- `[:punct:]` --- a collection of punctuation characters
- `[:symbol:]` --- a collection of symbols
- `[:digit:]` --- a digit
- `[:lower:]` --- same as `[a-z]`, a lower-case letter
- `[:upper:]` --- same as `[A-Z]`, an upper-case letter
- `[:alpha:]` --- standing for alphabet, same as `[a-zA-Z]`, `[:lower:]|[:upper]`
- `[:alnum:]` --- standing for alphanumeric, letter or number, `[:alpha:]|[:digit:]`
- `[:graph:]` --- anything but whitespace: punctuation, symbols, digits, letters
    - same as `[:punct:]|[:symbol:]|[:alnum:]`
    
## Parentheses and Capture Groups

* A potential shortcoming of the previous patterns described is we have no way to "self-reference".

* For example, `[^aeiou]{2}` matches any two consecutive consonants, which don't have to be the same.

* What if we need them to be the same, and we don't want to check for "bb" or "cc" or "dd"...?

> Surrounding a regular expression in (parentheses) allows you to reference it later with `\\#`, such as `\\1` for the first set of () or `\\2` for the second.

* To match the same consonant twice in a row:

```{r}
str_view(color_vector, "([^aeiou])\\1", match = NA)
```

* Which words have the same letter repeated but not consecutively? Match (and capture) any letter, then one or more "buffer" letters, and then a repeat of the first letter.

```{r}
str_view(color_vector, "(.).+\\1")
```

* Notice that "aliceblue" has another match which is not showing up here; "eblue".

* Also notice that "aquama" is only counted once in "aquamarine".

```{r}
colors_small %>% 
  mutate(matches = str_count(colors, "(.).+\\1"))
```

* These findings are examples of two general principles.

> Regular expressions are **greedy**; they will match the LONGEST POSSIBLE sequence that they can.

> When detecting occurrences of a pattern, only distinct, non-overlapping occurrences will count separately.

```{r}
str_view("rrr", "rr")
str_count("rrr", "rr")
```

* You can also use parentheses to just control the logic of a regular expression, even if you don't care about reusing the match.

* For example, let's revisit: which colors contain a vowel that is immediately followed by EITHER another vowel OR a digit?

```{r}
# Our first solution
str_view(color_vector, "[aeiou][aeiou0-9]", match = NA)
```

```{r}
# An attempt to reasonably use "or" which won't work; this will read as (vowel-vowel) or (digit)
str_view(color_vector, "[aeiou][aeiou]|[:digit:]", match = NA)
```

```{r}
# Parentheses make this work as vowel - (vowel or digit)
str_view(color_vector, "[aeiou]([aeiou]|[:digit:])", match = NA)
```



---

##### EXERCISE: Write Your Own Regular Expression

* This exercise will give you practice writing your own regular expressions.

*Note: These can be accomplished without regular expressions with relative ease. I ask you to use regular expressions as an exercise.*

* Our example data will be `fruit`, a vector of 80 common fruits.

```{r}
fruit
```

* **Write** the regular expression needed to answer the following questions.

* Which fruits contain the pattern "berry" or "melon"?

```{r}
# str_view(fruit, "your pattern here")
str_view(fruit, "berry|melon")

```

* Which fruits contain a "b", followed by exactly three characters followed by any character which is NOT y?

```{r}
# str_view(fruit, "your pattern here")
str_view(fruit, "b.{3}[^y]")
```


* Which fruits start AND end with a vowel?
    - *Hint: Your regular expression should translate to plain english as "start of string, then a vowel, then any number of any character, then a vowel, end of string"*

```{r}
# str_view(fruit, "your pattern here")
str_view(fruit, "^[aeiou].+[aeiou]$")
```

> Philosophical takeaway: **Take your time** building regular expressions. One character, one concept at a time. Rome wasn't built in a day, neither are regular expressions.

---    
    
## Matching Special Characters Literally With The Backslash

```{r}
ugly_equations = c("$22.00 + 2^3 = 30", "sqrt(-1) = ?")
```

* Let's say we wanted to detect which equations contain the literal symbol "+".

* A reasonable first attempt would be:

```{r, error = TRUE}
str_view(ugly_equations, "+")
```

* That ugly error is occurring because our regular expression has a syntax problem.

* Namely, `+` has special meaning in a regular expression, meaning repeat the previous symbol one or more times. We didn't give it a symbol to repeat one or more times.

* A reminder of how `+` behaves by default:

```{r}
# Match one or more digits consecutively
# Reminder: Regular expressions are greedy, so it will match two digit numbers as one expression, rather than the two digits separately
str_view(ugly_equations, "[:digit:]+")
```

> To match a special character **literally**, precede it with two backslashes: `\\`.

```{r}
str_view(ugly_equations, "\\+", match = NA)
```

* Which equations start with `$`?

```{r}
str_view(ugly_equations, "^\\$", match = NA)
```

* Which equations end with a `?`?

```{r}
str_view(ugly_equations, "\\?$", match = NA)
```

# Exoplanets

- The *exoplanet* data set has many string variables
- Let's return to and examine them using `stringr`.

## Read the Data

```{r}
planets = read_csv("../../data/exoplanet-confirmed_2022-09-12.csv",
                   skip = 124) %>%
  select(pl_name, hostname, discoverymethod, sy_pnum, disc_year) %>%
  rename(planet = pl_name,
         star = hostname,
         method = discoverymethod,
         number=sy_pnum,
         year = disc_year) %>%
  mutate(index = row_number()) %>% #Add an index variable
  select(index, everything())
head(planets)
```

## Last Character

> Find the last character of each planet name. 

```{r}
planet_letters = planets %>%
  mutate(letter = str_sub(planet,-1, -1))

planet_letters %>% count(letter)
```

> Produce a bar graph of how often each last character occurs.

```{r}
ggplot(planet_letters, aes(x=letter)) +
  geom_bar(color="black", fill="thistle") +
  labs(
    x = "Planet name suffix",
    y = "Count",
    title = "NASA Exoplanet Archive",
    subtitle = "As of June 15, 2021"
  )
```

* By default, these are appearing in alphabetical order.

* We now introduce the `reorder()` command to remedy this.

### Sidenote: reorder()

* In technical terms...

> `reorder(x, X, FUN = mean)` changes the variable `x` into a factor variable, with levels in order of ascending `FUN(X)` within each unique value of `x`.

* In plain English...

> `reorder(columnName, order_by)` is often used so that the values of `columnName` appear in order of their values of `order_by` on a plot.

* When you have only one row per mark, the choice of `FUN` is somewhat arbitrary and you can leave it as the default; if you have more than one row mapping to a mark on the graph, you should consider whether you are ordering by `mean`, `min`, `median`, `max`, or something else! (see `?reorder`)

```{r}
letter_counts = planet_letters %>% group_by(letter) %>% summarize(n = n())

ggplot(letter_counts, aes(x = reorder(letter, n), y = n)) +
  geom_col(color="black", fill="thistle") +
  labs(
      x = "Planet name suffix",
      y = "Count",
      title = "NASA Exoplanet Archive",
      subtitle = "As of June 15, 2021"
    )

ggplot(letter_counts, aes(x = reorder(letter, desc(n)), y = n)) +
  geom_col(color="black", fill="thistle") +
  labs(
      x = "Planet name suffix",
      y = "Count",
      title = "NASA Exoplanet Archive",
      subtitle = "As of June 15, 2021"
    )
```

---

## "c"'s Without "b"'s

> Are there any planets in the data set whose name ends in " c" for which there is not a corresponding planet in the same stellar system whose name ends " b"?

```{r}
##Find star names with planets that have the suffix c
star_c = planet_letters %>%
  filter(letter=="c")

##Find b planets
star_b = planet_letters %>%
  filter(letter=="b")

##Find stars in star_c but not in star_b
star_c %>%
  anti_join(star_b, by = join_by(star))
```

## Planet Naming Patterns

> Most planets have a name that matches the star name followed by a space and a lower case letter. Identify any planets that do not follow this pattern.

- The simple approach here can fail if the star name contains characters with special meaning in a regular expression.

```{r}
# Notice that this isn't matching when the star name has a +
planets %>%
  select(planet, star) %>%
  filter(!str_detect(planet, star)) #star name in planet
```

```{r}
# This addresses the + matching issue and returns planets that do not follow the convention
planets %>%
  select(index, planet, star) %>%
  # Set up a regular expression (notice the space after the star name):
  mutate(star_regex = str_c('^',star," ")) %>%
  #Use "\\\\+" to get the \\+ needed in the regular expression
  mutate(star_regex = str_replace_all(star_regex, "\\+", "\\\\+")) %>%
  #Remove planets with names that include the name of the host star:
  filter(!str_detect(planet, star_regex))
```

