---
title: "STAT 240: readr & lubridate"
author: "Cameron Jones"
date: "Spring 2024"
output: rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      error = TRUE, fig.height = 4)
library(tidyverse)
library(lubridate)
library(readxl)

## A script to change default colors in ggplot2
source("../../scripts/viridis.R")
```

# Overview

## Learning Outcomes

- These lectures will teach you how to:
    + Read in data from different file types
    + Convert dates in string format to Date objects
    + Extract components from dates
    
## Preliminaries

- You will need the `readxl` package for these lecture files, which you can obtain with `install.packages("packageName")`.

### Files

- Files to download to `COURSE/lecture/unit4-readr`
    - `week4-readr.Rmd`
    - `week4-readr.html`
  
- Files to download to `COURSE/data/`
  - `Obesity+Prevalence+by+ZIP+Code+and+Age+Group,+Females.xlsx`    
  - `Obesity+Prevalence+by+ZIP+Code+and+Age+Group,+Males.xlsx`
  
## Supplementary Material

- [*R4DS* Chapter 11](https://r4ds.had.co.nz/data-import.html) for information on **readr**
- [*R4DS* Chapter 16](https://r4ds.had.co.nz/dates-and-times.html) for information on **lubridate**

# readr - Importing Data

* **Reading** data into R is the process of extracting information from some external file and storing that information in a variable we can work with.

* So far this semester, we have always given you `.csv` files, given you code to read the data in using `read_csv()`, and focused on everything that happens after that.

* However, data does not always come in a friendly `.csv` format.

> Data can be stored in many different ways, and the function you use to read the data into R depends on the way it is stored.

* The file extension (such as `.csv`) of a data file often, but not always, tells you about its form.

## Delimiters

> Every data file has a **delimiter**, which indicates when an individual data units starts and ends.

* Delimiters are helpful both to the user and to the `read_*` commands.

* For example, consider the following sequence of names:

```
Matt Ryan Tom Brady
```

* This could be reasonably interpreted as four separate first names.

* It could also be reasonably interpreted as the two famous NFL quarterbacks, Matt Ryan and Tom Brady. However, a `read_*` command would never read it this way, because there is no (unique) delimeter between those two full names.

* Consider how you would expect a `read_*` command to interpret each of these:

```
Matt Ryan, Tom Brady
```

```
Matt; Ryan; Tom; Brady
```

```
Matt Ryan  Tom   Brady
```

* The last one in particular is an interesting case; if a single space is the delimiter, there is a blank field between "Ryan" and "Tom", and TWO blank fields between "Tom" and "Brady". Or, if ANY amount of space is a single delimiter, there are no blank fields.

* With the general function `read_delim()`, you can read in data with ANY delimiter.

* However, many common cases have better shortcut functions.

## Comma Separated Values

> In a `.csv` file, standing for Comma Separated Values, the comma ("`,`") serves as the delimiter between individual data values.

* `.csv` is the most common file type for storing data.

* A `.csv` file is a plain text document. There is nothing fancy going on within it, such as html tags or internal code. It is just text, including commas.

* The **first line** usually contains the column names, separated by commas.

* The following lines are the rows of the dataframe (as opposed to how you manually construct one in R, which is column by column), with each field separated by commas.

* We have been using `read_csv()` to read in information from `.csv` files; it will also accept a string with the above specifications.

```{r}
read_csv("
         column1, column2, column3
         T, row1val2, 24,
         F, row2val2, 25
         ")
```

* Notice that column1, column2, and column3 are the column names, not a row of the dataframe.

* Notice that column1 read in as "logical", column2 read in as "character", and column3 is "double" (numeric). 

* Many default behaviors of the `read_*` functions can be modified with arguments to suit the individual needs of a given file. We will explore some here.

### Guessing Column Types

* Let's mix up some strings and numbers to see how it handles each situation.

```{r}
read_csv("
         column1, column2, column3
         row1val1, 24, 24
         26, row2val2, 25
         ",
         )
```

* This example illustrates an important point of how `read_csv` decides upon column types.

* **All numbers can be treated as words, but not all words can be treated as numbers**, so if a column has even one non-digit (except the decimal point `.`) in it, the whole column will be treated as words.

* This is true for other data types as well; so many columns will "accidentally" be read in as character.

* You can override the column types that R chooses using the `col_types` argument, with a call to the `cols()` function within:

```{r}
read_csv("
         column1, column2, column3
         T, 24, 24
         F, row2val2, 25
         ",
         col_types = cols(
           column1 = col_character(),
           column3 = col_character()
           )
         )
```

* You can also just use `mutate()` and one of the `as.something` commands we've seen the past two weeks.

### Handling Unfriendly Files

* Sometimes, `.csv` files will have information in them that is NOT part of the data.

```{r}
read_csv("Here's some information about this file that I created, pretty neat huh?
         Here's a little more information. 
         column1, column2
         top left, top right
         bottom left, bottom right
         ")
```

* There was a comma in the first line, so `read_csv` broke it into two columns.

* There was no comma in the second line, so `read_csv` leaves an `NA` in the second column for the "first row", the one that contains "Here's a little more information."

* For this situation, you can tell `read_csv` to skip the first `n` lines with the `skip` argument.

```{r}
read_csv("Here's some information about this file that I created, pretty neat huh?
         Here's a little more information.
         column1, column2
         top left, top right
         bottom left, bottom right
         ",
         skip = 2)
```

* Some `.csv` files will be even meaner and not give you the column names.

```{r}
read_csv("Here's some information about this file that I created, pretty neat huh?
         The column names are in another castle!
         top left, top right
         bottom left, bottom right
         ",
         skip = 2)
```

* You can handle this by telling `read_csv` that `col_names = FALSE`, and it will treat the first row of data it reads as the first row of the dataframe, not the column names.

```{r}
read_csv("Here's some information about this file that I created, pretty neat huh?
         The column names are in another castle!
         top left, top right
         bottom left, bottom right
         ",
         skip = 2,
         col_names = FALSE)
```

* The columns are automatically named `X1`, `X2`, et cetera.

* All of the custom behaviors we have displayed using `read_csv` are applicable to the other `read_*` commands we are about to go through as well.

## read_csv2()

* Many countries around the world use the comma to delinate the start of decimal values in a number, whereas the typical U.S. symbol is the decimal point ("`.`"). 

* In this setting, the comma no longer works as a delimiter, because `123,45` represents one hundred twenty three and fourty five hundreths (`123.45`), NOT two separate numbers.

* Data files that use this convention typically use a semicolon ("`;`") as their delimiter.

> You can read in files which use `;` as the delimiter (typically due to `,` being the decimal indicator) with `read_csv2()`.

```{r}
# incorrectly trying to use read_csv
read_csv("someNumber; someText
          123,45; ABC
          81,66; DEF
         ")
```

```{r}
read_csv2("someNumber; someText
          123,45; ABC
          81,66; DEF
         ")
```


## read_tsv()

> Files where a tab is the delimiter can be read with `read_tsv()`; those files often have the extension `.tsv`.

* The tab is a non-constant amount of whitespace designed to keep columns vertically aligned in text files.

* In R, the tab can be represented with `\t`. This is called a **special character**, and we'll learn more about it next week in stringr.

```{r}
read_tsv("col1\tcol2\tcol3
          4\t8\t15
          16\t23\t42")
```

## read_table()

> Files where any whitespace is the delimiter can be read with `read_table()`. They often have the extension `.txt`.

* Notice how the amount of whitespace differs between each field; `read_table()` ignores how many spaces there are and just skips right to the next non-space character.

```{r}
read_table("column1 column2  column3
           value1     value2          value3")
```

## read_delim()

* The four cases above: `read_csv()`, `read_csv2()`, `read_tsv()`, and `read_table()` account for the common simple data file types.

* There are other more complicated data forms like the hierarchical JSON and HTML files, and there exist packages with commands to deal with those.

* However, you have the power to read any data file given that it uses one consistent delimiter, with `read_delim()`.

> If you encounter a data file with an uncommon delimiter, you can use `read_delim()`, the general form of all previous `read_*` commands.

* Here's an example where, for some reason, `+` is the delimiter.

```{r}
read_delim("column1+column2+column3
           4+8+15
           16+23+42",
           delim = "+")
```

* Just as we represented a tab with `\t`, we can represent a "newline" (or "return", or "enter") with `\n`.

```{r}
read_delim("column1+column2+column3\n4+8+15\n16+23+42",
           delim = "+")
```

## read_excel()

* Excel files are fundamentally different than the above file types. Because Excel cells can be merged and split in irregular ways, the definitions of columns and rows can get murky.

* However, the `readxl` package includes the function `read_excel` which deals with data like this intelligently; and can also helpfully read data with no such irregularities.

* Assuming you have installed the `readxl` package and loaded it with `library(readxl)` at the top of these lecture notes, you should be able to read in the following data.

* You should see a list of zip codes, starting with 53001, and some mostly missing obesity data.

* We will be using a cleaned version of these files next week.

```{r}
women = read_excel("../../data/Obesity+Prevalence+by+ZIP+Code+and+Age+Group,+Females.xlsx",
                   sheet = 3,
                   skip = 5,
                   na = ".")
men = read_excel("../../data/Obesity+Prevalence+by+ZIP+Code+and+Age+Group,+Males.xlsx",
                   sheet = 3,
                   skip = 5,
                   na = ".")

head(women)
head(men)
```

# lubridate - Date Parsing

* `Date` objects in R represent a specific day of a specific month of a specific year.

* From a Date object, we can extract out interesting features such as the day of the week it was, apply date-specific operations such as "add two weeks", and cross-reference it with other date-based information.

* However, R doesn't always recognize which information is meant to be dates, because they can be expressed in many different ways. 

* For example, the date "April 31st, 2001" could be written as:
    - 4/31/2001
    - 4-31-2001
    - 31 April 2001
    - April 31 2001
    - 4 31 2001
    - 31 4 2001
    - ...and many more unforeseen formats!
    
* There are two problems that arise:
    - Inconsistent formats (slashes, dashes, spaces) make it hard to write a simple conversion function.
    - Ambiguous dates (does 1/4 represent January 4th or April 1st?) may be incorrectly interpreted by R
    
* Often, you need to convert information that was read in as a string to be a Date type.

## Parsing a Date 

> The lubridate functions `dmy()`, `dym()`, `mdy()`, `myd()`, `ydm()`, and `ymd()`, take a provided string and convert it to a Date object, based on the **order of the components in the provided string**.

* The lubridate functions above are extremely flexible, and can handle almost any format of date.

* However, the ambiguous dates problem here means you must have knowledge of how the given date is formatted.

* The order of the components in the provided string determines what function you use.

* If the date is provided in the order **m**onth, **d**ay, **y**ear, you would use `mdy()`.
    - If it is in the order **y**ear, **m**onth, **d**ay, you would use `ymd()`, and so on.
    - All six possible orders have their own functions.

```{r}
date1 = "February 5 2017"
class(date1) # R doesn't recognize that this is a date

# The order is Month - Day - Year, so we use mdy() to convert it to a Date object
converted = mdy(date1)
converted # lubridate will SHOW you dates in the international standard format, year-month-day, but that is irrelevant to the function you pick
class(converted)
```

```{r date-from-string-examples}
date2 = "5-2-2017"
date3 = "2017, February 5"
date4 = "Feb. 5, 2017"
date5 = "02/05/2017"
date6 = "2.2017.5"
```

- `date2` could be read as two separate dates, either May 2, 2017 or February 5, 2017.

```{r date-from-string-examples2}
date2
dmy(date2)
mdy(date2)
```

* Once you have identified the correct function from the order of the components, the lubridate functions will handle the rest.

```{r date-from-string-examples3}
##
date3
ymd(date3)
##
date4
mdy(date4)
##
date5
mdy(date5)
##
date6
myd(date6)
```

---

##### EXERCISE: Date Parsing Practice

* This exercise will give you practice with basic lubridate functions.

* **Write** code in each chunk to convert each of the following practice date strings into a Date object using a lubridate function.

```{r}
# This date represents February 3rd, 2024
practiceDate1 = "2024-3-2"

# ???(practiceDate1)
```

```{r}
# This date represents May 4th, 2023
practiceDate2 = "5.4.24"

# ???(practiceDate2)
```

```{r}
# This date represents April 12th, 2002
practiceDate3 = "12/04/02"

# ???(practiceDate3)
```

---

## Extracting Components from a Date

- We can extract components of dates from a date-valued object.

```{r date-components}
test_date = mdy(date1)
test_date
```

- The function `year()` extracts the year.

```{r}
year(test_date)
```

- The function `month()` extracts the month. See variations if you want the number, an abbreviation of the month, or the full name.

```{r}
test_date
month(test_date)
month(test_date, label=TRUE)
month(test_date, label=TRUE, abbr=FALSE)
```

- `day()` extacts the day of the month

```{r}
test_date
day(test_date) # day of the month
```

- `wday()` extracts the day of the week.
  - note that Sunday is `1`
  - just like `month()`, you can get an abbreviated day of the week or the full day
  
```{r}
test_date
wday(test_date) # day of the week
wday(test_date, label=TRUE)
wday(test_date, label=TRUE, abbr=FALSE)
```

- The function `yday()` gets the day of the year.
- The value is different in the months March and later in leap years compared to non-leap years.

```{r}
# a leap year
march_01_2020 = ymd("2020-03-01")
march_01_2020
yday(march_01_2020)

# not a leap year
march_01_2022 = ymd("2022-03-01")
march_01_2022
yday(march_01_2022)
```

---

##### EXERCISE: Predict the Output

* This exercise will reinforce lubridate date parsing and component extraction.

* **Predict** what the following code will produce; then run it and see if you were right!

```{r}
day(mdy("January 30 24"))
```

```{r}
month(dym("1.2024.3"), label = TRUE, abbr = FALSE)
```

```{r}
year(ydm("02/03/04"))
```

---
