---
title: "STAT 240: Airport Case Study"
author: "Cameron Jones"
date: "Spring 2024"
output: rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      error = TRUE, fig.height = 4)
library(tidyverse)
library(scales)

## A script to change default colors in ggplot2
source("../../scripts/viridis.R")
```


# Read in Data

- We can now read in the data from six different airports and combine into a single data set
- Here, we do this with one line for each data set and then use `bind_rows()` to combine them.
- Then, we can delete the temporary files

```{r read-csv}
den = read_csv("../../data/DEN-2018.csv")
jfk = read_csv("../../data/JFK-2018.csv")
lax = read_csv("../../data/LAX-2018.csv")
msp = read_csv("../../data/MSP-2018.csv")
ord = read_csv("../../data/ORD-2018.csv")
sfo = read_csv("../../data/SFO-2018.csv")

awt = bind_rows(den, jfk, lax, msp, ord, sfo)
# awt is at the airport-day-hour level

rm(den, jfk, lax, msp, ord, sfo)
# rm standing for "remove", deletes the objects we no longer need to free up some computer memory
```

# Passengers by Terminal

> How many passengers and flights went through each airport-terminal combination (in 2018)? Display this in a summary table, with all the terminals from the most popular airports displayed first, and the next most popular airport's terminals shown second, and so on - with terminals sorted within airport by their number of passengers.

* Note that the `all_total` column represents total passengers within that terminal-day-hour, and `all_flights` represents total flights within that terminal-day-hour. We will need to summarize these columns to the level of airport-terminal.

- Then, **add a column** `total_airport_passengers` which includes the total number of passengers in the corresponding airport across all terminals. *(Note: this column will repeat values for airports with multiple terminals.)*

* Arrange the table by descending order of `total_airport_passengers` (this will sort the airports by most to least international flights) and then break ties by the total number of passengers in that terminal.

* Remove `total_airport_passengers` and print the table.

```{r}
## airports and terminals
awt %>%
  group_by(airport, terminal) %>%
  summarize(passengers = sum(all_total),
            flights = sum(all_flights)) %>% 
  group_by(airport) %>% 
  mutate(total_airport_passengers = sum(passengers)) %>% 
  arrange(desc(total_airport_passengers), desc(passengers)) %>% 
  select(-total_airport_passengers)

### JFK and LAX have five different international terminals
### The other airports have one or two each.
```

# Days of Data

> How many unique days of data are there for each airport?

* Problem: We can't just count how many times each airport occurs; because different airports have different numbers of terminals, and each terminal might not necessarily have data for every day-hour.

```{r}
awt %>% 
  count(airport)
```

```{r key-variables-airport-dates}
## examine date by airport
awt %>%
  group_by(airport, date) %>% 
  summarize() %>% # A shortcut for group_by(airport, date) %>% summarize() is distinct(group_by, summarize)
  group_by(airport) %>% 
  summarize(n = n())
## One date is missing for all the airports
```



# Which Date is Missing?

> Identify which date is missing for each airport; is it the same day for each airport?

- One day is missing for each airport!  Can we figure out which one it is?
- One approach could be to narrow down by:
    - adding a month variable
    - counting the number of distinct days within each month
    - seeing which total is less than the actual number of days
    - find the distinct days within that month
- But a more straightforward way as we did with the Madison weather data is to:
    - create a mini data frame with all possible dates
    - join with a data summary of the dates and terminals in the data set
    - see what is missing
    
```{r}
all_dates = tibble(
  date = seq(ymd("2018-01-01"), ymd("2018-12-31"), 1)
)

airport_dates = awt %>% distinct(airport, date) # returns all the unique combinations of airport & date which appear in awt

# return all the dates which appear in all_dates that don't appear in airport_dates
anti_join(all_dates, airport_dates)
```


```{r}
# Here's how you can manually verify it is the same missing date for each airport; the "rep" command stands for "replicate" or "repeat". We are creating a dataframe with 365*6 rows, one row for each airport-day combination.

## Create a mini temporary data set with all dates in 2018 for each airport code
temp1 = tibble(
  date = rep(seq(ymd("2018-01-01"), ymd("2018-12-31"), 1), times = 6),
  airport = rep(c("DEN", "JFK", "LAX", "MSP", "ORD", "SFO"), each = 365)
)

## Create another temporary data set with the actual dates and airports in awt

temp2 = awt %>% 
  select(airport, date) %>% 
  distinct()

## Use anti_join() to find rows in temp1 not in temp2
temp1 %>% 
  anti_join(temp2)
```

- We see that May 8, 2018, is missing from each airport. Curious!
- Verify this

```{r}
## check on May 8
awt %>%
  filter(date == "2018-05-08")

### No airports have any data on May 8, 2018.
```

# Checking Totals

> Does the `all_total` variable match the sum of the other columns of passenger counts?

- Add a column named `check_total` with the sum of the columns which names begin with "all" (except for `all_total`)
- Compare these totals with that in `all_total`

- Filter `awt` to produce a dataframe with the rows where `check_total`, the manually computed total, does not match `all_total`, the given values.

```{r}
## check that all_total column matches
awt = awt %>%
  mutate(check_total = all_n_0_15 +
            all_n_16_30 +
            all_n_31_45 +
            all_n_46_60 +
            all_n_61_90 +
            all_n_91_120 +
            all_n_120_plus +
            all_excluded)

table_mismatch = awt %>%
  filter(all_total != check_total)
table_mismatch
```

- There are 44 rows where the totals do not match

# Investigating Mismatches

>  At what time of day and at which airports are these mismatches happening?


```{r}
# Recall: We could have done this with count(airport, hour)
table_mismatch %>%
  group_by(airport, hour) %>% 
  summarize(n = n())
### All mismatches are between midnight and 1am.
### Most are at LAX and JFK
```

> Create a histogram to show the distribution of the size of these mismatches.

```{r}
ggplot(table_mismatch, aes(x = all_total - check_total)) +
  geom_histogram(boundary=0, binwidth=5, fill = "red", color = "black")
```

- In each of the 44 cases where there is a mismatch, the `all_total` variable exceeds the `check_total`
- This likely means that some of the passengers in the `all_total` column are not included in the specific waiting time counts of passengers.
- Most of the differences are fewer than five passengers
- The largest difference is somewhere between 30 and 35

# Busy Months

> Are some months busier than others?

- Begin by adding a month variable and then reordering the columns
- Also add day and weekday for future use

```{r}
awt = awt %>% 
  mutate(month = month(date, label = TRUE),
         day = day(date),
         wday = wday(date, label = TRUE)) %>% 
  relocate(month, day, wday, .after = hour)
```

- Total the number of passengers by month and sort by that total, descending.

```{r}
awt %>%
  group_by(month) %>%
  summarize(total = sum(all_total)) %>%
  arrange(desc(total))
```

- The previous summary is affected by different numbers of days in each month.
    - A better solution calculates an average number of daily passengers separately for each month
- Calculate the average daily rate of passengers by month.
- (Recall that May has only 30 observations.)

- Key idea here is to:
    - first summarize by date (and month and day):
    - sum to get daily totals across airport, hours, and terminal
    - then group by month
    - calculate means within the month

```{r}
busy_months = awt %>%
  group_by(date, month, day) %>%
  summarize(total = sum(all_total)) %>%
  group_by(month) %>%
  summarize(n = n(), total = sum(total)) %>%
  mutate(daily = total/n) %>% 
  arrange(desc(daily))

busy_months
```

> Display these average daily passengers by month in a bar graph.

```{r}
ggplot(busy_months, aes(x=month, y=daily)) +
  geom_col(fill = "blue", color = "black") +
  scale_y_continuous(labels = scales::comma) +
  xlab("") +
  ylab("# Daily Arrivals") +
  ggtitle("Daily International Arrival Totals", subtitle = "at six selected US Airports")
```

- It looks like the summer months (July, August, June) are the busiest in terms of number of passengers traveling per day for the airports in our data.

# Busy Weekdays

> Are some days of the week busier than others?  

```{r}
busy_wdays = awt %>%
  group_by(date, wday) %>%
  summarize(total = sum(all_total)) %>%
  group_by(wday) %>%
  summarize(n=n(), total = sum(total))  %>%
  mutate(daily=total/n) %>%
  arrange(desc(daily))

busy_wdays

ggplot(busy_wdays, aes(x = wday, y = daily)) +
  geom_col(fill = "blue", color = "black") +
  scale_y_continuous(labels = scales::comma) +
  xlab("Day of the week") +
  ylab("Average number of passengers") +
  ggtitle("Are some days of the week busier than others?")
```

# Busy Times of the Day

> Are some times during the day busier than others?  

```{r busy-times}
busy_times = awt %>%
  group_by(date, hour) %>%
  summarize(total = sum(all_total)) %>%
  group_by(hour) %>%
  summarize(n = n(), total = sum(total))  %>%
  mutate(hourly = round(total/n,2)) %>%
  arrange(desc(hourly)) 
busy_times

ggplot(busy_times, aes(x = hour, y = hourly)) +
  geom_col(fill = "blue", col = "black") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  +
  xlab("Time of the day") +
  ylab("# of passengers") +
  ggtitle("Average Hourly International Arrivals", subtitle = "at six select US airports")

```


# Airport Efficiency

## By Wait Time

> Are some airports more efficient than others in terms of average wait time? Show the distribution of each airport's `all_avg_wait` times with a boxplot.

```{r efficient}
ggplot(awt, aes(x=airport, y=all_avg_wait)) +
  geom_boxplot()
```

- We can adjust the plot by reordering the airports in order by the median of their average waiting time for all passengers.
- We also add better axis labels


```{r}
awt = awt %>% 
  mutate(airport = reorder(airport, all_avg_wait, FUN=median))
## treat airport as a factor with levels ordered according to the median of the all_avg_wait variable

ggplot(awt, aes(x=airport, y=all_avg_wait)) +
  geom_boxplot() +
  xlab("Airport") +
  ylab("Average waiting time for all passengers") +
  ggtitle("Are some airports more efficient that than others in terms of average wait time?")
```

- Passengers tend to wait longer on average at San Francisco's airport than the others in this analysis.

> What is the average waiting time at each airport for all passengers?

- To calculate this statistic, we need to sum the waiting times for all passengers and divided by the total number of passengers
- Note that it is incorrect to calculate the average for each airport/terminal/date/hour and then average these values as the number of passengers in each row is different.
  - We want to average by passenger, not by row in the data set
- The strategy is to calculate the total time by all passengers in each row
  - This is the average time multiplied by the number of passengers
- Then, sum times and passengers by airport and divide

```{r}
awt = awt %>% 
  mutate(total_time = all_total * all_avg_wait)

prob_2 = awt %>% 
  group_by(airport) %>% 
  summarize(avg_wait = sum(total_time) / sum(all_total))

prob_2 %>% 
  arrange(avg_wait)
```

- Looks like San Francisco is slower than the other airports

## By Passengers Processed Per Booth

> Do airports differ in the number of passengers processed per booth?

- You can filter out rows where `all_booths` is 0.
- Add a new column, `passengers_per_booth`, which is the ratio of the number of passengers (`all_total`) per booth (`all_booths`)
- For each airport, compute the minimum, median, and maximum of `passengers_per_booth`; calculate 
- Print this table, sorted by descending median of `passengers_per_booth` 

```{r}
awt = awt %>% 
  mutate(passengers_per_booth = all_total / all_booths)

prob_1 = awt %>% 
  filter(all_booths > 0) %>% 
  group_by(airport) %>% 
  summarize(min = min(passengers_per_booth),
            median = median(passengers_per_booth),
            max = max(passengers_per_booth))

prob_1 %>% 
  arrange(desc(median)) %>% 
  print(n = Inf, width = Inf)
```

- MSP has the largest number of passengers per booth
  - The previous graph showed MSP also had the shortest distribution of waiting times.

> Display the median values in a bar graph

```{r}
ggplot(prob_1, aes(x = airport, y = median)) +
  geom_col() +
  xlab("Airport") +
  ylab("Passengers per Booth") +
  ggtitle("Comparison of Median Passengers per Booth",
          subtitle = "International arrivals for six US Airports")
```

