---
title: "STAT 240: Airport Case Study"
author: "Cameron Jones"
date: "Spring 2024"
output: rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      error = TRUE, fig.height = 4)
library(tidyverse)
library(scales)

## A script to change default colors in ggplot2
source("../../scripts/viridis.R")
```

# Overview

## Learning Outcomes

- These lectures will reinforce dplyr concepts through interactive live coding practice on a real case study.
    
## Preliminaries

- You will need the `scales` package for these lecture files, which you can obtain with `install.packages("packageName")`.

## Files

- Files to download to `COURSE/lecture/unit4-readr`
    - `week4-airport.Rmd`
    - `week4-airport.html`
  
- Files to download to `COURSE/data/`
  - `DEN-2018.csv`
  - `JFK-2018.csv`
  - `LAX-2018.csv`
  - `MSP-2018.csv`
  - `ORD-2018.csv`
  - `SFO-2018.csv`
  
- Files to download to `COURSE/scripts`
  - `COURSE/scripts/viridis.R`
  
## Supplementary Material

- [*CNCS* Chapter 10](https://bookdown.org/bret_larget/stat-240-case-studies/airport-waiting-times.html) for background on the airport waiting time data

# The Airport Case Study: more dplyr

- For this analysis, we will read in international arrival data from several airports in the US.
- The data is collected by the US Customs and Border Control.
- Details about the data are described in [Chapter 11 of Course Notes and Case Studies](https://bookdown.org/bret_larget/stat-240-case-studies/airport-waiting-times.html).

- We will look at international flight arrival data from 2018 from six US Airports
  - DEN (Denver)
  - JFK (New York)
  - LAX (Los Angeles)
  - MSP (Minneapolis and St. Paul)
  - ORD (Chicago, O'Hare)
  - SFO (San Francisco)

- Each data file is at the **terminal-day-hour** level. E.g., for every day in the year 2018, there are up to 24 separate hours recorded in that day, for each terminal within that airport.

- They contain **terminal-day-hour** information about the number of arriving international flights and the number of passengers, among other information like customs waiting times.

- When we bind all of the **terminal-day-hour** airport files together into one, we end up with one large file at the level of **terminal-day-hour**.

## Read in Data

- We can now read in the data from six different airports and combine into a single data set
- Here, we do this with one line for each data set and then use `bind_rows()` to combine them.
- Then, we can delete the temporary files

```{r read-csv}
den = read_csv("../../data/DEN-2018.csv")
jfk = read_csv("../../data/JFK-2018.csv")
lax = read_csv("../../data/LAX-2018.csv")
msp = read_csv("../../data/MSP-2018.csv")
ord = read_csv("../../data/ORD-2018.csv")
sfo = read_csv("../../data/SFO-2018.csv")

awt = bind_rows(den, jfk, lax, msp, ord, sfo)
# awt is at the airport-day-hour level

rm(den, jfk, lax, msp, ord, sfo)
# rm standing for "remove", deletes the objects we no longer need to free up some computer memory
```

# Live Coding Questions

## Passengers by Terminal

> How many passengers and flights went through each airport-terminal combination (in 2018)? Display this in a summary table, with all the terminals from the most popular airports displayed first, and the next most popular airport's terminals shown second, and so on - with terminals sorted within airport by their number of passengers.

* Note that the `all_total` column represents total passengers within that terminal-day-hour, and `all_flights` represents total flights within that terminal-day-hour. We will need to **summarize** these columns to the level of airport-terminal.

- After that, we'll **add a column** `total_airport_passengers` which includes the total number of passengers in the corresponding airport across all terminals. *(Note: this column will repeat values for airports with multiple terminals.)*

* Arrange the table by descending order of `total_airport_passengers` (this will sort the airports by most to least international flights) and then break ties by the total number of passengers in that terminal.

* Remove `total_airport_passengers` and print the table.

```{r}

```

## Days of Data

> How many unique days of data are there for each airport?

* Problem: We can't just count how many times each airport occurs; because different airports have different numbers of terminals, and each terminal might not necessarily have data for every day-hour.

```{r}
awt %>% 
  count(airport)
```

```{r}
# Live coding
```

## Which Date is Missing?

- One day is missing for each airport!  Can we figure out which one it is?
- One approach could be to narrow down by:
    - adding a month variable
    - counting the number of distinct days within each month
    - seeing which total is less than the actual number of days
    - find the distinct days within that month
- But a more straightforward way as we did with the Madison weather data is to:
    - create a mini data frame with all possible dates
    - join with a data summary of the dates and terminals in the data set
    - see what is missing

> Identify which date is missing for each airport; is it the same day for each airport?

* We will need the `anti_join` function for this question; we will study this function in more depth next week.

* `anti_join(x, y)` returns all rows in a dataframe `x` which do NOT have a "match" in the dataframe `y`.

```{r}
# Live Coding
```

- We see that May 8, 2018, is missing from each airport. Curious!

```{r}
## check on May 8
awt %>%
  filter(date == ymd("2018-05-08"))
```

## Checking Totals

> Does the `all_total` variable match the sum of the other columns of passenger counts?

- Add a column named `check_total` with the sum of the columns which names begin with "all_n", and "all_excluded"
- Compare these totals with that in `all_total`

- Filter `awt` to produce a dataframe with the rows where `check_total`, the manually computed total, does not match `all_total`, the given values.

```{r}
# Live Coding
```

- There are 44 rows where the totals do not match

## Investigating Mismatches

>  At what time of day and at which airports are these mismatches happening?

```{r}
# Live Coding
```

> Create a histogram to show the distribution of the size of these mismatches.

```{r}
# Live Coding
```

- In each of the 44 cases where there is a mismatch, the `all_total` variable exceeds the `check_total`
- This likely means that some of the passengers in the `all_total` column are not included in the specific waiting time counts of passengers.
- Most of the differences are fewer than five passengers
- The largest difference is somewhere between 30 and 35

## Busy Months

> Are some months busier than others?

- Begin by adding a month variable to `awt` and then reordering the columns
- Also add day and weekday for future use

```{r}
# Live Coding
```


- Total up the number of passengers by month and sort by that total, descending.

```{r}
# Live Coding
```

- The previous summary is affected by different numbers of days in each month.
    - A better solution calculates an average number of daily passengers separately for each month
- Calculate the average daily rate of passengers by month.
- (Recall that May has only 30 observations.)

- Key idea here is to:
    - first summarize by date (and month and day):
    - sum to get daily totals across airport, hours, and terminal
    - then group by month
    - calculate means within the month

```{r}
# Live Coding
```

> Display these average daily passengers by month in a bar graph.

```{r}
# Live Coding
```

- It looks like the summer months (July, August, June) are the busiest in terms of number of passengers traveling per day for the airports in our data.

## Busy Weekdays

> Are some days of the week busier than others?  

```{r}
# Live Coding
```

## Busy Times of the Day

> Are some times during the day busier than others?  

```{r}
# Live Coding
```

## Airport Efficiency

### By Wait Time

> Are some airports more efficient than others in terms of average wait time? Show the distribution of each airport's `all_avg_wait` times with a boxplot.

```{r}
# Live Coding
```

- We can adjust the plot by reordering the airports in order by the median of their average waiting time for all passengers.

```{r}
# Live Coding
```

> What is the average waiting time at each airport for all passengers?

- To calculate this statistic, we need to sum the waiting times for all passengers and divided by the total number of passengers

- Note that it is incorrect to calculate the average for each airport/terminal/date/hour and then average these values as the number of passengers in each row is different.
  - We want to average by passenger, not by row in the data set
  
- The strategy is to calculate the total time by all passengers in each row
  - This is the average time multiplied by the number of passengers
  
- Then, sum times and passengers by airport and divide

```{r}
# Live Coding
```

### By Passengers Processed Per Booth

> Do airports differ in the number of passengers processed per booth?

- You can filter out rows where `all_booths` is 0.
- Add a new column, `passengers_per_booth`, which is the ratio of the number of passengers (`all_total`) per booth (`all_booths`)
- For each airport, compute the minimum, median, and maximum of `passengers_per_booth`; calculate 
- Print this table, sorted by descending median of `passengers_per_booth` 

```{r}
# Live Coding
```

- MSP has the largest number of passengers per booth
  - The previous graph showed MSP also had the shortest distribution of waiting times.

> Display the median values in a bar graph

```{r}
# Live Coding
```



