---
title: "STAT 240: dplyr Practice Solutions"
author: "Cameron Jones"
date: "Spring 2024"
output: rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,  message=FALSE, warning=FALSE,
                      error = TRUE,
                      fig.height = 4)
library(tidyverse)
library(lubridate)
```

# Read in Data

```{r}
## read in the data
official = read_csv("../../data/madison-weather-official-1869-2022.csv")
```

# Annual Average Temperature

> Create a summary table of the data with the average temperature per year. Then, make an effective plot of this data to show patterns over time and variation around the main pattern.

```{r}
mw1 = official %>% 
  ## drop any rows with missing data in tavg
  drop_na(tavg) %>% 
  ## use the lubridate function year() to extract the year from the date
  mutate(year = year(date)) %>% 
  ## group by year
  group_by(year) %>% 
  ## calculate the average temperature by each year
  summarize(yearlyTemp = mean(tavg))

mw1
```

```{r}
ggplot(mw1, aes(x = year, y = yearlyTemp)) +
  geom_point() +
  geom_line() +
  geom_smooth(se = FALSE) +
  labs(
    x = "Year",
    y = "Temperature (F)",
    title = "Madison Average Annual Temperature Over Time",
    subtitle = "1869-2021"
  )
```

# Daily Temperature Records

> For each of the 366 unique days of the year, find the historical date which had the highest maximum temperature on that day of the year. There may be ties, which will lead to slightly more than 366 rows.

```{r}
official %>% 
  mutate(year = year(date),
         month = month(date, label = TRUE),
         day = day(date)) %>% 
  group_by(month, day) %>% 
  slice_max(tavg, n = 1)
```

# Temperature by 30-year period

> Meteorologists often determine weather norms by averaging over a 30-year period. Create a variable which partitions the data into 30-year periods from 1871-1900, 1901-1930, up to 1991-2020. You can exclude years before 1871 and after 2020.

> Then calculate within each of the twelve months within each period (e.g. 1 row for all Januarys in 1871-1900, 1 row for all Februarys in 1871-1900...) the average, maximum, and minimum of the average temperature.

```{r}
# Create variables and filter
prob2 = official %>% 
  mutate(year = year(date),
         month = month(date, label = TRUE)) %>% 
  filter(year > 1870 & year < 2021) %>% 
  mutate(period30 = case_when(
    year > 1870 & year <= 1900 ~ "1871-1900",
    year > 1900 & year <= 1930 ~ "1901-1930",
    year > 1930 & year <= 1960 ~ "1931-1960",
    year > 1960 & year <= 1990 ~ "1961-1990",
    year > 1990 & year <= 2020 ~ "1991-2020"))  

prob2 %>% count(period30) # appears to have worked correctly
```

```{r}
# Summarize to the month-period level
prob2_summarized = prob2 %>% 
  # Drop those 14 dates with missing tavg
  drop_na(tavg) %>% 
  group_by(month, period30) %>% 
  summarize(minimum = min(tavg), maximum = max(tavg), average = mean(tavg))

prob2_summarized
```

# Days With Precipitation

> Make a summary table which shows the proportion of days which had any precipitation in each 30-year period + month combination.

```{r}
# prob2 is the same as official, at the date level, with period30 and month already created

prob4 = prob2 %>% 
  drop_na(prcp) %>% 
  group_by(month, period30) %>% 
  summarize(any_prcp = sum(prcp > 0), 
            all_days = n(), 
            percentPrcp = any_prcp/n())

prob4
  
```

> With a separate panel for each month, create a column graph of the percentage of precipitation in each 30-year period.

```{r}
ggplot(prob4, aes(percentPrcp, period30)) +
  geom_col() +
  facet_wrap(~month)
```





