---
title: "Discussion 2 "
author: "Cameron Jones"
date: 'Week of September 19'
output: html_document
---

```{r, echo = FALSE, message = FALSE}
library(tidyverse)

Orange$Tree = factor(Orange$Tree, levels = c("1", "2", "3", "4", "5"))
```

# Logistics

- Discussion 1, Homework 1 assignments graded
  + Take a look at grades/comments
  + Let me know if you have any questions
  
- In discussion groups for 5+ weeks; get to know each other!
  + A group-mate may be on Zoom
  + Even if you are all in-person, it is nice to share your screen with everyone else so you don't crowd one laptop
  
- Discussion assignments
  + Only **one person** has to submit.
  + You should submit at the end of the discussion period **even if you didn't finish**.
  

# Content Review: ggplot2

#### Our Data: Orange Tree Circumference vs. Age

```{r}
Orange[c(1:3, 8:10),] # "Orange" is one of many of R's built-in datasets! You can see them all using data().
```

# Concept: ggplot() Arguments and Aesthetics

```{r, message = FALSE, fig.height = 3, fig.width = 6}
ggplot(data = Orange, mapping = aes(x = age, y = circumference)) + 
  
  # Start every plot with ggplot(data, aesthetics)
  # Aesthetics which are listed here, called "global" aesthetics, will apply to any geometry functions added below
  
  geom_point()

```

#### Variable vs. Constant Aesthetics

```{r, message = FALSE, fig.height = 3, fig.width = 6}
ggplot(data = Orange, mapping = aes(x = age, y = circumference)) + 
  
  # If I want to change a property of the geometry on the plot, (e.g. its color or size) in a constant way across the entire plot (e.g. make every point red and large) we do NOT list those in aes().
  
  geom_point(color = "red", size = 5)
```

```{r, message = FALSE, fig.height = 3, fig.width = 6}
ggplot(data = Orange, mapping = aes(x = age, y = circumference, color = Tree)) + 
  
  # If I want to change a property of the geometry based on a VARIABLE IN THE DATASET (e.g. color each point by the individual tree identifier), that is done by aes().
  
  geom_point(size = 5) + 
  geom_line()

  # ggplot automatically creates the appropriate legend when you specify a variable aesthetic.
```

# Concept: Faceting

```{r}
ggplot(data = Orange, mapping = aes(x = age, y = circumference)) +
  geom_point() +
  geom_line() +
  
  # In some scenarios, it may be useful to plot each tree on its own plot, rather than all of them together. facet_wrap() takes in a single CATEGORICAL variable, splits the dataset into multiple datasets based on the levels of that categorical variable, and plots them all separately.
  
  facet_wrap(vars(Tree))

# facet_wrap will do its best to choose a reasonable number of rows and columns.

```

```{r}
ggplot(data = Orange, mapping = aes(x = age, y = circumference)) +
  geom_point() +
  geom_line() +
  
  facet_grid(. ~ Tree)

# facet_grid(rows ~ cols) allows you to specify alternate layouts
```


```{r}
ggplot(data = Orange, mapping = aes(x = age, y = circumference)) +
  geom_point() +
  geom_line() +
  
  facet_grid(Tree ~ .)
```

# Concept: Residuals

#### What is a residual?

```{r}
x = data.frame(x = c(1, 2, 3), y = c(3, 7.5, 8), yhat = 2*c(1,2,3) + 2)

ggplot(x, aes(x, y)) +
  geom_point(size = 5) +
  geom_line(aes(x, yhat), color = "blue")
```

```{r}
ggplot(x, aes(x, y)) +
  xlim(1, 3) +
  ylim(2, 8) +
  geom_point(size = 5) +
  geom_line(aes(x, yhat), color = "blue") +
  
  annotate("segment", x = 1, xend = 1, y = 4, yend = 3, color = "red") +
  annotate("segment", x = 2, xend = 2, y = 6, yend = 7.5, color = "red") +
  annotate("segment", x = 2.86, xend = 2.99, y = 6.5, yend = 7.87, arrow = arrow(ends = "last", length = unit(.2, "cm"))) +
  annotate("text", x = 1.1, y = 3.5, label = "r = -1", color = "red") +
  annotate("text", x = 2.1, y = 7, label = "r = 1.5", color = "red") +
  annotate("text", x = 2.86, y = 6.4, label = "r = 0", color = "red")
```

#### Example: Bad Fit

```{r}
Orange$residualsBad = mean(Orange$circumference) - Orange$circumference

ggplot(Orange, aes(age, circumference, size = abs(residualsBad))) +
  geom_point() +
  geom_hline(yintercept = mean(Orange$circumference), color = "blue") +
  theme(
    legend.position = "none"
  )
```

```{r}
ggplot(Orange, aes(x = residualsBad)) +
  geom_histogram(fill = "papayawhip", color = "black", bins = 8) +
  xlim(c(-100, 100)) +
  ylim(c(0, 15))
```

```{r}
ggplot(Orange, aes(x = residualsBad)) +
  geom_density(fill = "papayawhip")
```

#### Example: The Best Fit "Regression" Line

```{r, message = FALSE}
model = lm(circumference ~ age, Orange)

Orange$residuals = model$residuals

ggplot(Orange, aes(age, circumference, size = abs(residuals))) +
  geom_point() +
  geom_smooth(method = "lm", se = F) +
  scale_size_continuous(limits = c(0, 100)) +
  theme(
    legend.position = "none"
  )
```

```{r}
ggplot(Orange, aes(x = residuals)) +
  geom_histogram(fill = "papayawhip", color = "black", bins = 8) +
  xlim(c(-100, 100))
```

```{r}
ggplot(Orange, aes(x = residuals)) +
  geom_density(fill = "papayawhip") +
  xlim(c(-100, 100))
```
