---
title: "Discussion 3"
author: "Cameron Jones"
date: 'Week of September 26'
output: html_document
---

```{r, echo = FALSE, message = FALSE}
library(tidyverse)
```

# Concept Review: dplyr

The built-in R dataframe "chickwts" contains weight measurements on 71 baby chickens after being fed one of six diets for a month and a half. 

```{r}
slice_sample(chickwts, n = 8)
```

# Concept: Filtering Rows and Columns with filter(), slice_max(), drop_na(), and select()


## filter()

filter() reduces the dataframe to the set of ROWS which satisfy a given condition.

```{r}
chickwts %>% 
  filter(feed == "horsebean") # Double equals '==' tests for equality.

chickwts %>% 
  filter(weight <= 140)

# Other (in)equality testers: <, <=, >=, >, %in% (coming up!)
```

## slice_max()

slice_max() returns a specified number of rows (n) which have the highest values in the dataframe of a certain variable. slice_min() returns the lowest.

```{r}
chickwts %>% 
  slice_max(weight, n = 5)

chickwts %>%
  slice_min(weight, n = 2)
```


## drop_na()

drop_na() removes the rows which have an NA (missing data) in them.

```{r}

# Adding rows with NA's for testing.

chickwts[72,] = c(NA, "sunflower")
chickwts[73,] = c(300, NA)

tail(chickwts)


nrow(chickwts) # 73 rows in the dataset

chickwts %>% 
  drop_na() %>% # Removes any row with an NA anywhere
  nrow() 

chickwts %>% 
  drop_na(feed) %>% # Removes any row with an NA in the "feed" column
  nrow() #

```

```{r, echo = FALSE}
rm(list=ls())
```

## select()

select() reduces a dataframe to a set of specified COLUMNS.

```{r}
chickwts %>% 
  select(feed) %>%
  head()
```


# Concept: Creating Summary Tables with group_by() and summarize()

group_by() and summarize() work together to reduce a dataframe of many rows down to a small dataframe which contains summary measures.

```{r}

# Summarize alone, with no groups, just returns a specified summary of the entire dataframe

chickwts %>% 
  summarize(avgWeight = mean(weight),
            stdDevWeight = sd(weight))

# When a group_by is specified before summarize, those summaries are provided within each group

chickwts %>% 
  group_by(feed) %>% 
  summarize(avgWeight = mean(weight),
            stdDevWeight = sd(weight))  

```

# Concept: The %in% operator

```{r}

# The %in% operator tests if a given value is contained within a specified vector.

"phrase to find" %in% c("not here", "phrase to find", "not here either")

7.5 %in% 1:10

chickwts %>% 
  filter(feed %in% c("linseed", "soybean", "cheez-its")) %>% 
  count(feed)
```

