---
title: "Discussion 6"
author: "Cameron Jones"
date: 'Week of October 17'
output: html_document
---

# Logistics

* We are working hard to grade your midterms
* Turning point in the course
  +   Up to now: R/tidyverse basics, "data manipulation" crash course
  +   From here on out: More statistical/mathematical concepts, *implementing* and *reporting* on these concepts in R
  +   Your final project (eventually) will need you to demonstrate skills at both!
* Announcement on Canvas: **OPTIONAL, UNGRADED, ANONYMOUS** Discussion Feedback Survey (coming soon)
  +   An anonymous opportunity to tell me how I'm doing as a TA, how you are benefitting or not benefitting from discussion
* Starting next week: new discussion groups!

#### Regular Expression Resources

- https://evoldyn.gitlab.io/evomics-2018/ref-sheets/R_strings.pdf
- regex101.com

```{r, echo = FALSE, message = FALSE}
library(tidyverse)
```

# Concept Review: Stringr

### Data: Fruit Strings

```{r}
fruit # fruit is one of stringr's built in dataframes! It contains 80 strings with the names of fruits.
```

## Concept: The "detection" str_functions, str_detect & str_count

These commands *detect* or *count* occurrences of a given pattern. They don't return strings, they return TRUE/FALSE and a number, respectively.

#### str_detect

```{r, error = TRUE}
#str_detect(string(s) to check, pattern to check for)

str_detect("blueberry", "berry") # TRUE because blueberry contains berry

str_detect(c("blueberry", "orange"), "berry") # TRUE, FALSE; it operates on EACH word, returning a TRUE or FALSE for each word if you pass in a list

str_detect(fruit, "berry") %>% sum() # Recall; the sum of a TRUE/FALSE list is the number of TRUEs! 14 of the 80 fruits in 'fruit' have the word "berry" in them.
```

#### str_count

```{r}
#str_count(string(s) to check, pattern to check for)

str_count("blueberry", "b") # Returns 2 because there are 2 b's in "blueberry"

str_count(c("blueberry", "orange"), "b") # Returns a list, (2, 0)

str_count(fruit, "an") %>% max() # One of our fruits has the pattern "an" twice; which fruit?
```

## Concept: The "extraction" str_functions, str_sub & str_extract

These commands "extract" certain parts of strings; i.e. the first three letters of a string, or the part of a string which follows a particular pattern. They return STRINGS.

#### str_sub

```{r}

#str_sub(string(s) to check, start position, end position)
# Think of it as standing for "subsection"

str_sub("blueberry", 2, 3) # The second character to fourth character (inclusive) in the string
str_sub("blueberry", 5, 5) # The fifth character alone
str_sub(c("blueberry", "orange"), 4, -2) # The fourth to second-last character in the strings

```

#### str_extract

```{r}

#str_extract(string(s) to check, pattern to extract)

str_extract("blueberry", "berry") # Returns "berry" as the part of "blueberry" which matches the pattern
str_extract(c("blueberry", "orange"), "berry") # Returns (berry, NA) as the parts of each string which match the pattern

# This command is much more useful with regular expressions: see below

```


## Concept: Regular Expressions

How do I detect or extract a *complex* pattern, such as "at the very beginning of the string, a vowel followed by a consonant?"

We use the above commands in addition to a *regular expression*, which is a language which exists within most programming languages that allows for the concise and systematic specification of complex string patterns.

How many fruits START with a vowel, then a consonant?

```{r}
small = head(fruit)
small

str_detect(small, "^[aeiou][^aeiou]") # ^ refers to the beginning of the string, [aeiou] matches a character in the set {a, e, i, o, u}, and [^aeiou] matches a character NOT in the set {a, e, i, o, u}

str_extract(small, "^[aeiou][^aeiou]") # Again, recall the difference between count and extract


str_detect(fruit, "^[aeiou][^aeiou]") %>% sum() # 8 fruits start with a vowel and then a consonant.

# Which ones?

as_tibble(fruit) %>% filter(str_detect(fruit, "^[aeiou][^aeiou]")) # Recall that str_detect returns a list of TRUEs and FALSEs that we can pass into filter! (though filter only works on dataframes/tibbles, not a single list like fruit is)
```



