---
title: "Discussion 7 - Demo"
author: "Cameron Jones"
date: "Week of October 24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
source("../../scripts/ggprob.R")
```

# Logistics

+ HW5, Discussion 5 grades posted

+ In-person midterm grades posted; Take-home grades coming soon

+ New discussion groups starting today!


# Concept Review

### The Binomial Distribution

**Recall: What is a random variable?**
A random variable is a number which could take on a new value every time we run an experiment.

When we know certain things about the way a random variable is generated, we can assign probabilities to the different possible values that random variable might take on. The probabilities of the different values make up the **distribution** of the random variable.

One such example is the binomial distribution, which arises when our random variable is the number of "successes" from trials with:

+ *Binary outcomes*: A single trial either fails or succeeds.
+ *Independence*: Trials do not affect each other.
+ *Fixed number*: The number of trials is pre-determined.
+  *Fixed success probability*: Each trial has the same probability of succeeding.

For example; I know that if I flip a coin, I have a 50% chance of getting heads and a 50% chance of getting tails. Flip a fair coin twice; let $X$ be the number of heads I get. Because $X$ is the number of successes of a fixed number of independent trials, we know that $X$ follows the *"binomial distribution"*. With this knowledge, we can calculate how likely each possible value of $X$ is. There is a 1/4 chance I'll get 0 heads, 1/2 chance I'll get 1 head, and a 1/4 chance I'll get 2 heads.

```{r}
gbinom(n=2, p=0.5, size=3) +
    theme(axis.text=element_text(size=20),
        axis.title=element_text(size=22,face="bold"))
```

These probabilities are easily computed by hand, but what if I flip a fair coin 100 times? Knowing it follows the binomial distribution allows us to quickly produce all of the probabilities to the values which $X$ can take on.

```{r}
gbinom(n=100, p=0.5, size=3) +
    theme(axis.text=element_text(size=20),
        axis.title=element_text(size=22,face="bold"))
```

### Expectation and Variance

Knowing a variable follows a certain distribution is also useful because statisticians have derived shorter, more convenient formulas for the expectation and variance of these distributions.

For example, the general formula for the expectation (the average expected value) of a random variable is:

$$
\mu = \mathbb{E}(X) = \sum x P(X=x)
$$

However, when we know a variable follows the binomial distribution, this formula simplifies to:

$$
\mu = \mathbb{E}(Binom(n, p)) = n*p
$$

A similar result holds for the variance (a measure of the spread) of a random variable. The general formula is:

$$
\sigma^2 = Var(X) = \sum (x - \mu)^2 P(X=x)
$$

While the binomial special formula is:

$$
\sigma^2 = Var(Binom(n,p)) = n*p*(1-p)
$$

### Calculating Probabilities

The *pbinom(x, n, p)* command gives us the probability of getting $x$ or fewer successes from $n$ trials with probability of success $p$.

For example, the probability of getting 1 or fewer heads from 2 coin flips:

```{r}
pbinom(1, 2, 0.5)
```
This is the probability of getting 0 heads plus the probability of getting 1 head. *dbinom(x, n, p)* gives us the probability of getting exactly x successes.

```{r}
dbinom(0, 2, 0.5) + dbinom(1, 2, 0.5)
```
Now consider again the experiment of 100 coin flips. What is the probability of getting GREATER THAN 55 flips?

We know that:

$$
P(0 \leq X \leq 100) = 1
$$

Now consider the decomposition:

$$
P(0 \leq X \leq 100) = P(X \leq 55) + P(X > 55)
$$

Using the first fact and subtracting,

$$
P(X > 55) = 1 - P(X \leq 55)
$$

We arrive at this form, which is in terms that we can code with!

```{r}
1 - pbinom(55, 100, 0.5)
```
Graphically:

```{r}
gbinom(n=100, p=0.5, size=3) +
    theme(axis.text=element_text(size=20),
        axis.title=element_text(size=22,face="bold")) +
  geom_binom_density(100, 0.5, a = 55, b = 100, color = "red", size = 2.2) +
  geom_vline(xintercept = 55, color = "black", lty = "dashed", size = 2)
```



The red area is equal to the whole area (1) minus the blue area (P(X less than or equal to 55) = pbinom(55, 100, 0.5)).