---
title: "Discussion 8 - Demo"
author: "Cameron Jones"
date: 'Week of October 31'
output: html_document
---

```{r, message = FALSE, warning = FALSE, echo = FALSE}
library(tidyverse)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```

# Logistics

- HW6 Graded, HW7 Due Saturday

- Final Project Starts Today
  +   20 minutes discussion assignment, rest of time for project proposal
  +   Go to course home page, "Final Project Information" link, everything under "Project Proposal" due next Monday, Nov 7

## Concept Review

### Simulation vs. Theory

In this course, we often use _simulation_ to verify that observed results fall in line with theory.

Thus, the distinction between finding moments like the mean and standard deviation from _observed, simulated_ data and from _theoretical formulas_ is important to understand.

Let's say I hypothesize that the mean of a binomial variable is $n*p$. I can test this by generating random numbers from the binomial distribution and checking if the mean I observe is equal to n*p.

In this example, I test if the mean of $Binom(10, 0.5)$ is $10*0.5 = 5$.

```{r, echo = FALSE}
set.seed(4)
```


```{r}
numberOfSimulations = 3
n = 10
p = 0.5

# A single simulation; generate 3 values from the distribution Binom(10, 0.5)
simulation = rbinom(numberOfSimulations, n, p)
print(simulation)

# The mean of this simulation; is this n*p = 10*0.5 = 5?
print(simulation %>% mean() %>% round(4))

# If we do it again, we get a different value.
print(rbinom(numberOfSimulations, n, p) %>% mean() %>% round(4))

# How do we make sure the mean we get from the simulation is close to the actual mean? We sample a huge amount of random numbers! This should be closer to 5
numberOfSimulations = 100
print(rbinom(numberOfSimulations, n, p) %>% mean() %>% round(4))

# This should be almost exactly 5
numberOfSimulations = 1000000
print(rbinom(numberOfSimulations, n, p) %>% mean() %>% round(4))

```

The simulation verifies that the mean of $Binom(10, 0.5)$ is $10*0.5 = 5$.

### Binomial vs. Normal

The binomial distribution is _discrete_ ; it counts the number of successful trials, so it must be a non-negative integer. Recall the BINS acronym from lecture and last discussion for identifying binomial random variables.

```{r}
gbinom(10, 0.5, size = 2) +
  geom_binom_density(10, 0.5, a = 4, b = 4, color = "red", size = 2)
```

The probability that a binomial variable takes on an exact value is known and easily calculable; we have the closed mathematical formula for this, but all you need to know is that it is calculable with ```dbinom(x, n, p)```.

```{r}
dbinom(4, 10, 0.5) # The height of the red bar above; this is P(X = 2), where X ~ Binom(5, 0.6)
```

The normal distribution is _continuous_; it can take on any value in a range, not limited to whole or positive numbers.

```{r}
gnorm(5, 2, size = 2) +
  geom_segment(aes(x = 4, xend = 4, y = 0, yend = dnorm(4, mean = 5, sd = 2)), color = "red", size = 2)
```

The probability that a _normal_ variable takes on a specific value is _arbitrary_, because it is continuous. There are infinitely many real numbers in ANY interval, so the probability you get EXACTLY a certain value from a continuous distribution is effectively 0.

Attempting to calculate this with ```dnorm``` refers to the height of the density function at a specific point.

```{r}
dnorm(4, mean = 5, sd = 2) # The height of the red bar above.
```

```{r}
rnorm(100, 5, 2) 

# Notice how none of these random values are exactly 4, even though "pnorm tells us" a quarter of them should be
sum(rnorm(100, 5, 2) == 4)
```

Where as the probability of specific points has little meaning, _area_ under the normal curve does have real meaning.

```{r}
gnorm(5, 2, size = 2) +
  geom_norm_fill(5, 2, b = 4, fill = "red")
```

```{r}
pnorm(4, 5, 2) #The red area above; this is equivalent to P(X <= 4)!
```

