---
title: "Discussion 3"
author: "Cameron Jones"
date: 'Week of September 25'
output: html_document
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
```

# Concept Review: dplyr

#### Data: String Tension

The built-in tidyverse dataframe ```warpbreaks``` contains data on how many breaks appeared in a set length of yarn when exposed to a certain amount of tension.

```{r}
slice_sample(warpbreaks, n = 8)
```

## Concept: The Pipe (%>%) Operator

(helpful tip: keyboard shortcut for the pipe)

* Windows: ctrl+shift+M
* Mac: command+shift+M

```{r}
# What it does: Just passes the object on the left into the function on the "right" as the first argument
x = c(3, 4, 5, NA)

# These two lines of code are exactly equivalent
mean(x, na.rm = T)
x %>% mean(na.rm = T) # na.rm = T here is the SECOND argument, x is the first.
```

```{r}
# Why does it exist? Motivation: Making your dplyr work flow easier to read and produce!

# These two groups of code are exactly equivalent; dplyr commands take a dataframe in as the first argument and spit a dataframe out, so they can be connected with the pipe
warpbreaks %>% 
  group_by(tension) %>% 
  summarize(avgBreaks = mean(breaks)) %>% 
  arrange(desc(avgBreaks)) %>% 
  mutate(rank = row_number())

# Everything you can do with the pipe, you can do without it... but it's ugly and difficult to write
mutate(arrange(summarize(group_by(warpbreaks, tension), avgBreaks = mean(breaks)), desc(avgBreaks)), rank = row_number())
```

# Concept: Creating Summary Tables with group_by() and summarize()

group_by() and summarize() work together to reduce a dataframe of many rows down to a small dataframe which contains summary measures.

```{r, message = FALSE}
# group_by alone just adds an "instruction" to the dataframe that future mutates, summarizes, slices, etc. should use these groups, does not edit the dataframe at all. Notice that Groups: wool [2] has been added to the below output.
warpbreaks %>% 
  group_by(wool)

# Summarize alone, with no groups, just returns a single-row summary of the entire dataframe
# Note the syntax within summarize: newColumnName = formula(existingColumns), newColumn2 = formula(existingColumns)
warpbreaks %>% 
  summarize(avgBreaks = mean(breaks),
            sdBreaks = sd(breaks)) 

# When a group_by is specified before summarize, those summaries are provided within each group, defined by the categorical variable(s) you grouped by

warpbreaks %>% 
  group_by(wool) %>% 
  summarize(avgBreaks = mean(breaks),
            sdBreaks = sd(breaks)) 

# You can also group by multiple variables, in which case you will get one row for every unique combination of those two variables that appears

warpbreaks %>% 
  group_by(wool, tension) %>% 
  summarize(avgBreaks = mean(breaks),
            sdBreaks = sd(breaks)) 
```

