---
title: "Discussion 4"
author: "Cameron Jones"
date: 'Week of October 2'
output: html_document
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(lubridate)
```

# Concept Review: lubridate and dplyr II

### Data: Lakers 2008-2009

```{r}
# lakers is one of lubridate's built in dataframes! See them all with data().
lakers %>%
  select(date, etype, team, player, result, points) %>% 
  head()
```

## Concept: Parse Non-Dates into Dates with lubridate

When you read in data with dates in it, very often it will read in as a NUMBER (integer) or a WORD (string). 

However, if we want to do date-specific operations on it (e.g. extract the month or the day of the week), it needs to be a DATE.

```{r}
lakers %>% 
  distinct(date) %>% 
  head(n = 5)
```

This column is numeric right now. It is not reading 20081028 as October 28th, 2008; it is reading it as the number 20 million, 81 thousand and 28, so we cannot do date operations on it, such as getting the month.

```{r, error = TRUE}
lakers %>% 
  mutate(extractedMonth = month(date))
```

So how do we convert it to a date? The ORIGINAL date as we read it in, is in the form **Y**ear (2008) **M**onth (10) **D**ay (28), so we use the function **ymd()**.

Functions like ymd() exist for every possible form of dates, including mdy(), dmy(), et cetera. They are based on how the ORIGINAL date comes in.

```{r}
lakers2 = lakers %>% 
  rename(old_date = date) %>% 
  
  # change date from an INTEGER, in the form year-month-day, to a DATE
  mutate(new_date = ymd(old_date)) %>% 
  
  select(new_date, old_date)

lakers2 %>% 
  distinct() %>% 
  head()

str(lakers2)
```

Note the type of new_date is Date and the type of old_date is int.

## Concept: Extract Components from Dates

Once we have converted the column into a date, we can do date-based operations on it.

```{r}
lakers2 %>% 
  mutate(
    new_month = month(new_date, label = TRUE),
    new_weekday = wday(new_date, label = TRUE),
    new_dayInYear = yday(new_date)
  ) %>% 
  distinct() %>% 
  head()
         
```

## Complicated Dplyr Examples

Here are some extra practice dplyr problems. I display the final solution in the html, but NOT the solution code. **The solution code is in the Rmd.**

**Question:** Calculate the percent of shots the Lakers made in each game. Display the top three games by shot percentage.

+ Note: When I say "shots", I refer to rows where etype == "shot".

We have three tasks here:

+ filter to events which are shots taken by the Lakers (not their opponents)
+ calculate the proportion of made shots per game (reducing the dataframe to one row per game) (this is the tricky one for most students)
+ obtain the three highest games by shot percentage

```{r, echo = FALSE}
# Step 1
lakers_shots = lakers %>% 
  filter(etype == "shot" & team == "LAL")
```

```{r, echo = FALSE}
# Step 2
lakers_shots_calculations = lakers_shots %>% 
  
  # we want a row for each DATE, which identifies games uniquely
  # note that we don't have to convert this to a date; you can if you want, but we have no need to do any date-based operations on it
  group_by(date) %>%
  
  # Compute the following metrics within each DATE (game):
  summarize(
    # the expression result == "made" is a list of TRUE and FALSE. summing this list counts the number of TRUEs; i.e. the number of shots made
    n_shots_made = sum(result == "made"), 
    
    # within each DATE, how many rows are there? because we have filtered down to just shots, this is the number of shots taken
    n_shots_taken = n(),
    
    shot_percentage = n_shots_made/n_shots_taken
  )

# lakers_shots_calculations now has a row for every game, with the percentage of shots they made in that game
```

```{r, echo = FALSE}
# Step 3
lakers_shots_calculations %>% 
  slice_max(shot_percentage, n = 3)
```

**Question:** This dataset contains games in six different months, November through April. *Within each of those months*, which Lakers player had the highest free throw percentage? Players must have taken at least 20 free throws within that month to be included.

+ Note: Here we are looking at etype == "free throw".

```{r, echo = FALSE, message = FALSE}
lakers %>%
  # Step 1: Look at just Lakers free throws
  filter(etype == "free throw", team == "LAL") %>% 
  # Step 2: Create the "month" variable that we need
  mutate(actual_date = ymd(date),
         month = month(actual_date, label = TRUE)) %>% 
  
  # Within each player-month:
  group_by(player, month) %>% 
  summarize(
    # How many did they make?
    ft_made = sum(result == "made"),
    # How many did they take total? (number of rows within that player-location, because every row is a free throw)
    ft_taken = n(),
    # Divide the two for their percentage
    ft_percentage = ft_made/n()
  ) %>% 
  #ft_taken does not exist in the initial dataframe, it is created by the summarize above
  filter(ft_taken > 20) %>% 
  
  # Here we exploit the fact that slice_max has to obey the group_by function, and return the top 1 row (player) within each level of group_by (month)
  group_by(month) %>% 
  slice_max(ft_percentage, n = 1)
```


