---
title: "Simple Linear Regression in R"
author: "Ryan Yee"
date: "2023-04-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Libraries

```{r}
library(tidyverse)
```

## Dataset

We will use the built-in `iris` dataset which has 5 variables

```{r}
data("iris")
head(iris)
```

## SLR Model

We will fit a simipl linear regression in the following form:

$$ \text{Sepal.Width} = \text{Sepal.Length} * \beta + \alpha$$

where $\beta$ is our slope term and $\alpha$ is our constant or intercept term.
Note we are trying to predict Sepal Width from Sepal Length.

```{r}
slr <- lm(Sepal.Width ~ Sepal.Length, data = iris)
slr
```

Our estimates are $\beta = -0.06188$ and $\alpha = 3.41895$

## Model Summary

We can view a more useful summary using the `summary` function.

```{r}
summary(slr)
```

## Diagnostic Plots

We will need to look at diagnostic plots to justify the assumptions of our model are met. (This may or may not be discussed in lecture, so I will not go into details on how to assess the assumptions here).

```{r}
plot(slr)
```

## Visualize Model

We can visualize the model using `ggplot`.

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```



