---
title: "Simple Linear Regression in R"
author: "Ryan Yee"
date: "2023-04-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE}
library(tidyverse)
```

## Dataset

We will use FiveThirtyEight's Soccer Power Index dataset.
I am taking the first 1000 observations, there are over 11,000 observations in the full dataset (this will make it easier to visualize).
There are two key variables: `spi` which rates a team's strength and `xg` which is the number of goals we would expect a team to score in a game given the shots they took.

```{r, include=FALSE}
spi <- read.csv("../../../../projects/soccer/spi_matches_latest.csv") %>%
  select(spi1, spi2, xg1, xg2) %>%
  head(1000)
```

```{r}
head(spi)
```

```{r}
dim(spi)
```


## SLR Model

We will fit a simiple linear regression in the following form:

$$ \text{xG Difference} = \text{SPI Difference} * \beta + \alpha$$

where $\beta$ is our slope term and $\alpha$ is our constant or intercept term.
Note we are trying to predict the difference in expected goals from the difference in SPI or team strength.

```{r}
spi <- spi %>%
  mutate(
    xg_diff = xg1 - xg2,
    spi_diff = spi1 - spi2
  )

slr <- lm(xg_diff ~ spi_diff, data = spi)
slr
```

Our estimates are $\hat{\beta} = 0.04013$ and $\hat{\alpha} = 0.30159$

## Model Summary

We can view a more useful summary using the `summary` function.
This is useful for the interpretation portion of the analysis.

```{r}
summary(slr)
```

## Diagnostic Plots

We will need to look at diagnostic plots to justify the assumptions of our model are met. (This may or may not be discussed in lecture, so I will not go into details on how to assess the assumptions here).

```{r}
plot(slr)
```

## Visualize Model

We can visualize the model using `ggplot`.

```{r, warning=FALSE}
ggplot(spi, aes(x = spi_diff, y = xg_diff)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```



